Chasing \rainbows and \oceanglint:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-07 src/tex/ms.tex
[showyourwork] overleaf sync

\documentclass[
    usenatbib,
]{mnras}

\usepackage[T1]{fontenc}
\usepackage[dvipsnames, x11names]{xcolor}

@@

% Prettier fonts 
% (We might have to go back to defaults for MNRAS submission)
% \usepackage{libertine}
% \usepackage[libertine]{newtxmath}
% \usepackage[scaled=0.76]{beramono}

% Those seem to be the "right" fonts for MNRAS
\usepackage{newtxtext}
\usepackage{newtxmath}

\usepackage{microtype}
\usepackage{csquotes}
\usepackage[capitalise]{cleveref}
%\usepackage{xspace}

\usepackage{standalone}
\usepackage{tikz}
\usetikzlibrary{
    calc,
    positioning,
    shapes,
}

@@

%REMOVE LATER
\usepackage{soul}

@@

% Define macros
\newcommand{\IWA}{\ensuremath{\mathrm{IWA}}}
\newcommand{\OWA}{\ensuremath{\mathrm{OWA}}}
\newcommand{\hwo}{HabWorlds}

@@

\newcommand{\rainbows}{% MAK: KEEP this formatting for MNRAS submission. It deserves this!
    \textcolor{C0}{r}%
    \textcolor{C1}{a}%
    \textcolor{C2}{i}%
    \textcolor{C3}{n}%
    \textcolor{C4}{b}%
    \textcolor{C5}{o}%
    \textcolor{C6}{w}%
    \textcolor{C7}{s}%
    % \xspace
}

@@

% Chasing rainbows with the Habitable Worlds Observatory
\title{Chasing \rainbows{} and ocean glint:\\  Inner working angle constraints for the Habitable Worlds Observatory}

%phase angle probed by directly imaged planets, or
%Don't ask us about contrast

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan\thanks{Correspondence:~\url{sophia.vaughan@physics.ox.ac.uk}}\textsuperscript{,\afflink{1}},
    Kimberly Bott\textsuperscript{\afflink{2},\afflink{3},\afflink{4}},
    Sarah L. Casewell\textsuperscript{\afflink{5}},
    Nicolas B. Cowan\textsuperscript{\afflink{6}},
    David S. Doelman\textsuperscript{\afflink{7},\afflink{8}},
    \newauthor
    Timothy D. Gebhard\textsuperscript{\afflink{9},\afflink{10}},
    Matthew A.  Kenworthy\textsuperscript{\afflink{7}},
    Johan Mazoyer\textsuperscript{\afflink{11}},
    Maxwell A. Millar-Blanchaer\textsuperscript{\afflink{12}},
    \newauthor
    Olivier Absil\textsuperscript{\afflink{13}},
    Lisa Altinier\textsuperscript{\afflink{14}},
    Pierre Baudoz\textsuperscript{\afflink{11}},
    Ruslan Belikov\textsuperscript{\afflink{15}},
    Alexis Bidot\textsuperscript{\afflink{16}},
    Markus J. Bonse\textsuperscript{\afflink{10}},
    \newauthor
    Bernhard Brandl\textsuperscript{\afflink{7}},
    Alexis Carlotti\textsuperscript{\afflink{16}},
    Elodie Choquet\textsuperscript{\afflink{14}},
    Niyati Desai\textsuperscript{\afflink{17}},
    Kevin Fogarty\textsuperscript{\afflink{15}},
    Jules Fowler\textsuperscript{\afflink{18}},
    \newauthor
    Yann Gutierrez\textsuperscript{\afflink{11},\afflink{19},\afflink{20}},
    Olivier Guyon\textsuperscript{\afflink{21},\afflink{22},\afflink{23},\afflink{24}},
    Sebastiaan Y. Haffert\textsuperscript{\afflink{21}},
    Olivier Herscovici-Schiller\textsuperscript{\afflink{19}},
    \newauthor
    Adrien Hours\textsuperscript{\afflink{16}},
    Roser Juanola-Parramon\textsuperscript{\afflink{25},\afflink{26}},
    Elina Kleisioti\textsuperscript{\afflink{7}},
    Lorenzo König\textsuperscript{\afflink{13}},
    Mariya Krasteva\textsuperscript{\afflink{27}},
    \newauthor
    Iva Laginja\textsuperscript{\afflink{11}},
    Rico Landman\textsuperscript{\afflink{7}},
    Lucie Leboulleux\textsuperscript{\afflink{16}},
    David Mouillet\textsuperscript{\afflink{16}},
    Mamadou N’Diaye\textsuperscript{\afflink{28}},
    \newauthor
    Emiel H. Por\textsuperscript{\afflink{29}},
    Laurent Pueyo\textsuperscript{\afflink{29}},
    Frans Snik\textsuperscript{\afflink{7}},
    Daphne M. Stam\textsuperscript{\afflink{30}},
    Victor Trees\textsuperscript{\afflink{31},\afflink{32}},
    Dirk van Dam\textsuperscript{\afflink{7}},
    \newauthor
    Kyle van Gorkom\textsuperscript{\afflink{21}},
    Maaike van Kooten\textsuperscript{\afflink{33}}
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight that is reflected by an exoplanet can reveal angular features that are due to ocean glint, and to rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \hwo's coronagraph may obscure the phase angles where such characteristic features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
Planets in close-to-edge-on orbits have accessible phase angles limited by coronagraphic obscuration. 
%
We use \hwo'~catalog of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering by gas. 
%
We find that any exo-Earths in systems in this catalog will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The ocean glint signature at planetary scattering angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to \hwo~in $\sim\num{14}$ systems, and the rainbow due to water clouds at phase angles of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} \IWA{} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with the planet's orbital eccentricity and scales approximately inversely with the \IWA{}.  \end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%--------------------------------------------------------------------------------
%\begin{figure*}
    %\centering
    %\includegraphics[width=\linewidth]%{figures/Fig1IWAPol.png}
%    \caption{
%        The phase curves of two homogeneous planets with oceans and Earth-like atmospheres in edge-on orbits (the planets thus exhibit scattering angles 
%        ranging from 0$^\circ$ to 180$^\circ$ along their orbits). 
%        The cloudy scenario includes a low-altitude, Earth-like water 
%        cloud deck. 
%        Below the plot, we are showing the scattering angle ranges that cover the
%        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles (note that features partly overlap
%        with each other).  
%        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
%    }
%    \label{fig:bottplot}
%\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized polarized flux of a planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit (the planet thus exhibits phase angles ranging from \qty{0}{\degree} to \qty{180}{\degree} along its orbit). 
        The wind speed over the ocean is \qty{7}{\meter\per\second}. 
        The clouds in the planet's atmosphere consist of spherical water droplets. 
        The cloud coverage fraction is 0.5. 
        The cloud pattern is patchy, and the glint is kept cloud-free to maximize the glint signature \citep[see also Fig. 9 of][from which we took the mean signal of 300 planets with randomly generated patchy cloud patterns + 2 standard deviations, representative for cloud-free glint scenarios; the \enquote{noise} at the larger phase angles and longer wavelengths is due to the variation in the location of cloud patches]{treesandstam2019}. 
        The polarized flux is normalized to the total (unpolarized + polarized) flux (it is thus in fact the degree of polarization). 
        Note that the reference plane for the polarized flux is chosen such that only the polarized components perpendicular or parallel to the plane differ significantly from zero.
 %   and the latter
 %   is equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$.
        %The cloudy scenario includes a low-altitude, Earth-like water 
        %cloud deck. 
        Below the plot, we show the phase angle ranges that cover the
        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles (note that features partly overlap
        with each other).  
        %Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}

@@

The field of exoplanet research has made great advances in the last few decades. The characterisation of Earth-like planets in the habitable zones of their stars is now at the frontier of what we can achieve. The Habitable Worlds Observatory will perform a dedicated direct imaging survey of nearby stars searching for rocky planets in the habitable zone. This space-based coronographic telescope will detect and characterise exoplanets using their reflection spectrum. Complementary information on the habitability of these worlds can be obtained by observing the change in the planet's brightness at different planetary phases.

Variations of the flux and in particular the degree of polarization of
light of the parent star that is reflected by an exoplanet as it moves
along its orbit that are due to an ocean glint or optical phenomena from scattering by condensates --- like the rainbow --- could be some of the most distinct signatures of habitability.
%
However, depending on the architecture of a planetary system, some of these features can be obscured by the inner working angle (\IWA{}) of a telescope's coronagraph.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and, for example, geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.
%
Providing additional context to the habitability of an exoplanet would be the presence of a water cycle, as evidenced by the identification of a glint, i.e.\ the specular reflection of incident light by a liquid surface, or by a rainbow and other optical phenomena that can be attributed to light that is scattered by water clouds, as detailed in the next sections.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Exoplanet phase curves}

An exoplanet in an edge-on orbit moves through phase angles ranging from $0^\circ$ (full phase), through $90^\circ$ (quadrature), and to $180^\circ$ (the planet's nightside is towards the observer) (cf.\ Fig.~\ref{fig:bottplot}), while planets in higher inclination orbits will pass through a subset of these phase angles.
%
If the planet is rotating, its reflected light signal in both total and polarized flux is expected to show temporal variations that can place constraints on the horizontal heterogeneity of the planet \citep{2001Natur.412..885F,stam2008,2022A&A...664A..59M}.

The strengths of detailed total and polarized flux phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface, 2022A&A...664A..59M}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scattering, rainbows, and glories.
%
An added potential benefit of observing the state of polarization of the reflected light is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade as the light of these stars themselves can be considered to be unpolarized \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces. While glint is a familiar phenomenon on Earth, there is only
one other body in the Solar System where it has been observed: 
the Cassini orbiter captured the glint of sunlight reflecting off the
liquid hydrocarbon lakes near Titan's north pole as it was observing
the moon from the nightside \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%
The maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at an angle of reflection of \qty{53}{\degree} \citep[i.e., at a planetary phase angle of \qty{106}{\degree}; see, e.g.,][]{Zugger_2010,treesandstam2019}.
%
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
%

In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The stronger glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
The width across phase angles of the glint feature in total and polarized flux generally
broadens with increasing wave height \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}, which depends strongly on the wind speed over the surface, as described by \citet{CoxMunk1954}.
%

The glint provides evidence of specular reflecting surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
This confusion can be solved by measuring the intersection of phase curves at various wavelengths in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction \citep{treesandstam2019}, as shown in Fig.~\ref{fig:bottplot}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent, i.e.\ large phase angles, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry, as the glint changes the difference in polarization between continuum wavelengths and absorption bands. This could be applied when the planet is near quadrature, and thus more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, cloudbows, glories and other optical phenomena from light that is 
scattered by atmospheric condensates can dramatically influence the shape of a planet's total flux and polarized phase curves and provide a wealth of information 
about exoplanets \citep{karalidi2012rainbow, stam2008,Bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. Cloudbows refer to features occurring at the same phase angles from smaller cloud particles with the spread in colors invisible to the human eye making them appear white \citep[see][for further discussion]{Bailey2007}.  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the polarized phase curve of Venus at three wavelengths showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}.
%
The glory feature in Venus's total flux phase curve was later used to rule out potential solutions for the unknown UV absorber \citep{petrova2018glory}.

>>>>

\documentclass[usenatbib]{mnras}

\usepackage[T1]{fontenc}
\usepackage[dvipsnames, x11names, svgnames]{xcolor}

@@

% Prettier fonts 
% Those seem to be the "right" fonts for MNRAS
\usepackage{newtxtext}
\usepackage{newtxmath}

\usepackage{microtype}
\usepackage{csquotes}
\usepackage[capitalise]{cleveref}
\usepackage{xspace}

\usepackage{standalone}
\usepackage{tikz}
\usetikzlibrary{
    calc,
    positioning,
    shapes,
    shapes.symbols,
}

@@

@@

% Define macros
\newcommand{\IWA}{\ensuremath{\mathrm{IWA}}\xspace}
\newcommand{\OWA}{\ensuremath{\mathrm{OWA}}\xspace}
\newcommand{\HWO}{HabWorlds\xspace}

@@

\newcommand{\rainbows}{% MAK: KEEP this formatting for MNRAS submission. It deserves this!
    \textcolor{C0}{r}%
    \textcolor{C1}{a}%
    \textcolor{C2}{i}%
    \textcolor{C3}{n}%
    \textcolor{C4}{b}%
    \textcolor{C5}{o}%
    \textcolor{C6}{w}%
    \textcolor{C7}{s}%
    \xspace
}

\newcommand{\istar}{%
  \hspace{-0.125em}%
  \tikz[baseline=(char.base)]{
    \node [inner sep=0] (char) {i};
    \node[starburst, starburst points=5, starburst point height=2.5mm, fill=White, draw=MidnightBlue, line width=0.1pt, minimum width=0.3mm, minimum height=0.1mm, inner sep=0.9pt, rotate=15] at (0.0015, 0.15) {};
  }%
  \hspace{-0.175em}%
}
\newcommand{\oceanglint}{\textcolor{MidnightBlue}{ocean gl\istar{}nt}\xspace}

@@

% Full title of the paper (MNRAS has no running title?)
\title{Chasing \rainbows and \oceanglint:\\ Inner working angle constraints for the Habitable Worlds Observatory}

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan\thanks{Correspondence:~\url{sophia.vaughan@physics.ox.ac.uk}}\textsuperscript{,\afflink{1}},
    Kimberly Bott\textsuperscript{\afflink{2},\afflink{3},\afflink{4}},
    Sarah L. Casewell\textsuperscript{\afflink{5}},
    Nicolas B. Cowan\textsuperscript{\afflink{6}},
    David S. Doelman\textsuperscript{\afflink{7},\afflink{8}},
    \newauthor
    Timothy D. Gebhard\textsuperscript{\afflink{9},\afflink{10}},
    Matthew Kenworthy\textsuperscript{\afflink{7}},
    Johan Mazoyer\textsuperscript{\afflink{11}},
    Maxwell A. Millar-Blanchaer\textsuperscript{\afflink{12}},
    \newauthor
    Olivier Absil\textsuperscript{\afflink{13}},
    Lisa Altinier\textsuperscript{\afflink{14}},
    Pierre Baudoz\textsuperscript{\afflink{11}},
    Ruslan Belikov\textsuperscript{\afflink{15}},
    Alexis Bidot\textsuperscript{\afflink{16}},
    Jayne L. Birkby\textsuperscript{\afflink{1}},
    \newauthor
    Markus J. Bonse\textsuperscript{\afflink{10}},
    Bernhard Brandl\textsuperscript{\afflink{7}},
    Alexis Carlotti\textsuperscript{\afflink{16}},
    Elodie Choquet\textsuperscript{\afflink{14}},
    Niyati Desai\textsuperscript{\afflink{17}},
    \newauthor
    Kevin Fogarty\textsuperscript{\afflink{15}},
    Jules Fowler\textsuperscript{\afflink{18}},
    Yann Gutierrez\textsuperscript{\afflink{11},\afflink{19},\afflink{20}},
    Olivier Guyon\textsuperscript{\afflink{21},\afflink{22},\afflink{23},\afflink{24}},
    Sebastiaan Y. Haffert\textsuperscript{\afflink{21}},
    \newauthor
    Olivier Herscovici-Schiller\textsuperscript{\afflink{19}},
    Adrien Hours\textsuperscript{\afflink{16}},
    Roser Juanola-Parramon\textsuperscript{\afflink{25},\afflink{26}},
    Evangelia Kleisioti\textsuperscript{\afflink{7}},
    \newauthor
    Lorenzo König\textsuperscript{\afflink{13}},
    Mariya Krasteva\textsuperscript{\afflink{27}},
    Iva Laginja\textsuperscript{\afflink{11}},
    Rico Landman\textsuperscript{\afflink{7}},
    Lucie Leboulleux\textsuperscript{\afflink{16}},
    \newauthor
    David Mouillet\textsuperscript{\afflink{16}},
    Mamadou N’Diaye\textsuperscript{\afflink{28}},
    Emiel H. Por\textsuperscript{\afflink{29}},
    Laurent Pueyo\textsuperscript{\afflink{29}},
    Frans Snik\textsuperscript{\afflink{7}},
    \newauthor
    Daphne M. Stam\textsuperscript{\afflink{30}},
    Victor Trees\textsuperscript{\afflink{31},\afflink{32}},
    Dirk van Dam\textsuperscript{\afflink{7}},
    Kyle van Gorkom\textsuperscript{\afflink{21}},
    Maaike van Kooten\textsuperscript{\afflink{33}}
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows, and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \HWO's coronagraph may obscure the phase angles where such scattering features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in nearly edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that any exo-Earths in systems in this catalogue will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The rainbow due to water clouds at scattering angles of \qtyrange{140}{160}{\degree} would be accessible to \HWO\ in $\sim\num{42}$ systems, while ocean glint signature at scattering angles of $\qtyrange{50}{70}{\degree}$ would be accessible in $\sim\num{14}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.  %and scales approximately inversely with the \IWA.  
\end{abstract}

@@

@@

%\begin{figure*}
    %\centering
    %\includegraphics[width=\linewidth]%{figures/Fig1IWAPol.png}
%    \caption{
%        The phase curves of two homogeneous planets with oceans and Earth-like atmospheres in edge-on orbits (the planets thus exhibit scattering angles 
%        ranging from 0$^\circ$ to 180$^\circ$ along their orbits). 
%        The cloudy scenario includes a low-altitude, Earth-like water 
%        cloud deck. 
%        Below the plot, we are showing the scattering angle ranges that cover the
%        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles (note that features partly overlap
%        with each other).  
%        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
%    }
%    \label{fig:bottplot}
%\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized polarized flux of a planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit (the planet thus exhibits phase angles ranging from \qty{0}{\degree} to \qty{180}{\degree} along its orbit). 
        The wind speed over the ocean is \qty{7}{\meter\per\second}. 
        The clouds in the planet's atmosphere consist of spherical water droplets. 
        The cloud coverage fraction is 0.5. 
        The cloud pattern is patchy, and the glint is kept cloud-free to maximize the glint signature \citep[see also Fig. 9 of][from which we took the mean signal of 300 planets with randomly generated patchy cloud patterns + 2 standard deviations, representative for cloud-free glint scenarios; the \enquote{noise} at the larger phase angles and longer wavelengths is due to the variation in the location of cloud patches]{treesandstam2019}. 
        The polarized flux is normalized to the total (unpolarized + polarized) flux (it is thus in fact the degree of polarization). 
%        Note that the reference plane for the polarized flux is chosen such that only the polarized components perpendicular or parallel to the plane differ significantly from zero.
 %   and the latter
 %   is equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$.
        %The cloudy scenario includes a low-altitude, Earth-like water 
        %cloud deck. 
        Below the plot, we show the phase angle ranges that cover the
        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles adopted in this paper. Note that the features in the actual curves partly overlap
        with each other.  
        %Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}

@@

%The field of exoplanet research has made great advances in the last few decades. 
%The characterisation of Earth-like planets in the habitable zones of their stars is now at the frontier of what we can achieve. 
The Habitable Worlds Observatory (\HWO) will perform a dedicated direct imaging survey of nearby stars, searching for rocky planets in the habitable zone. 
This space-based coronagraphic telescope will detect and characterise exoplanets using their reflection spectrum. 
Complementary information on the habitability of these worlds can be obtained by observing the change in the planet's brightness at different planetary phases.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.

Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal the most distinct signatures of habitability: 
specular reflection (a.k.a. glint) from water oceans, or optical phenomena from scattering by condensates---like the rainbow.
%
However, depending on the architecture of a planetary system, some of these features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph.
%
%Providing additional context to the habitability of an exoplanet would be the presence of a water cycle, as evidenced by the identification of a  as detailed in the next sections.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Exoplanet phase curves}

%Exoplanets exhibit phase variations as they orbit their host star because

%
%If the planet is rotating, its reflected light signal in both total and polarized flux is expected to show temporal variations that can place constraints on the horizontal heterogeneity of the planet \citep{2001Natur.412..885F,stam2008,2022A&A...664A..59M}.

%The strengths of detailed total and polarized flux phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface, 2022A&A...664A..59M}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scattering, rainbows, and glories.
%
%An added potential benefit of observing the state of polarization of the reflected light is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade as the light of these stars themselves can be considered to be unpolarized \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off a smooth liquid surface. 
While glint is a familiar phenomenon on Earth, there is only one other body in the Solar System where it has been observed: the Cassini orbiter captured the glint of sunlight reflecting off the liquid hydrocarbon lakes near Titan's North Pole as it was observing the moon at crescent phases \citep{2010GeoRL..37.7104S}.
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
As shown in Figure \ref{fig:bottplot}, the maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at a scattering angle of \qty{53}{\degree} \citep[i.e., at a planetary phase angle of \qty{106}{\degree}; see, e.g.,][]{2008Icar..195..927W,Zugger_2010,treesandstam2019}.
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
\looseness=-1

In general, the glint on a planet will be strongest at red wavelengths because atmospheric Rayleigh scattering and attenuation are smallest \citep{Zugger_2011}. The stronger glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. %A degeneracy between polar brightness and glint is also smaller at redder wavelengths \citep{2012ApJ...752L...3C}.
%
The width across phase angles of the glint feature in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}, which depends strongly on the wind speed over the surface, as described by \citet{CoxMunk1954}.
%

Glint provides evidence of specular reflecting surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which motivates probing the rainbow angle for context, as described below.
%
This confusion can be resolved by measuring polarized phase curves at multiple wavelengths %in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction 
\citep{treesandstam2019}, as shown in \cref{fig:bottplot}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
%While the glint feature can only be probed when a planet is at crescent (i.e., large phase angles), \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry, as the glint changes the difference in polarization between continuum wavelengths and absorption bands. 
%This could be applied when the planet is near quadrature, and thus more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, cloudbows, glories and other optical phenomena from light that is scattered by atmospheric condensates can dramatically influence the shape of a planet's total flux and polarized phase curves and provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,Bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. 
    Cloudbows refer to features occurring at the same phase angles from smaller cloud particles, with the spread in colours invisible to the human eye making them appear white \citep[see][for further discussion]{Bailey2007}.  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the polarized phase curve of Venus at three wavelengths showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}.
%
The glory feature in Venus's total flux phase curve was later used to rule out potential solutions for the unknown UV absorber \citep{petrova2018glory}.
